Storm surges in the Strait of Georgia simulated with a regional model

### 2015-06-10 AO_guides/manuscriptAO.tex
Finished Keith's comments

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. In the future, storm surge forecasts will make use of Neah Bay water level predictions from the National Oceanic and Atmospheric Administration (NOAA).}

@@

{\color{red} The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. Further, a geostrophic slope calculation with velocity 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. Non-uniform forcing conditions will be investigated in a future study.}

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error. Note that the model will contain tidal energy from shallow water constituents due to nonlinear interactions in the model equations and so the correction as calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (Sa, Ssa, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model sea surface height as a correction. Note that the harmonic analysis used to generate the constituents is applied in the same manner as that used in the forcing, i.e.\, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy before the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported in terms of statistical measures of the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.  {\color{red} Observed and modelled maximum water levels and surge values as well as differences in their timing are reported in Table \ref{tab:peak}.}

@@

In general, there is also very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). {\color{red} The most noticeable difference is a broader peak in the model's surge curve, leading to a delay of a few hours in timing of the peak surge for most of the stations (Table \ref{tab:peak}). Note that the surge manifests as a broad peak in both the modelled and observed residual leading to a high probability that large residuals coincide with the high tide.} As such, it is of minor consequence that the timing of the observed and modelled maximum residuals do not match perfectly.

@@

\tbl{{\color{red}Summary of maximum water level, maximum surge amplitude and model delay for all simulations. The model time delay is the number of hours between the maximum in the observations and the maximum in the model. A negative number means the model's maximum was later than the observations. Values calucated over the same periods discusses in Table \ref{tab:statistics}. } }

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. This drop is smaller than the IB sea level change calculated by equation (\ref{eq:inverse}), approximately 1 cm rise in sea level associated with 1 hPa drop in atmospheric pressure. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This magnitude is in line with the findings of \citet{murty1995storm} that the IB contributes up to one third of the total surge amplitude. In our model, including the influence of remote atmospheric pressure forcing (in the sea level at the open boundary) produces realistic surge amplitudes compared with the runs only including local atmospheric forcing. Our results suggest that the local atmospheric pressure generates a sea level change that is smaller than the IB response. {\color{red} It is the atmopsheric pressure gradient that drives changes in the sea surface height. Since our domain is relatively small, so is the atmospheric pressure gradient, and hence, we see only a small contribution from local atmospheric pressure. }

@@

In a simulation without atmospheric forcing, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side (Figure \ref{fig:spatial_sshonly}). While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland {\color{red}(Figure \ref{fig:spatial})}.  {\color{red} Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps because the incoming surge wave could be partially reflected from the mainland coast in the Strait of Georgia. Since local winds are stronger in this case, their contribution to the surge amplitude of about 6 cm at Point Atkionson is more significant than in the Feb 2006 case. However, this is still less than 10 \% of the total surge amplitude. }

@@

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind events on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was {\color{red}71.8 cm} which is slightly higher than that observed at Point Atkinson {\color{red}(70 cm)}. Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. The model results do not reflect a higher surge at Victoria {\color{red}(54 cm versus 58.7~cm at Point Atkinson)}, perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.  {\color{red} There is a fairly significant delay in the timing of the maximum surge at most stations (Table \ref{tab:peak}). The model's surge curve is much broader and does not contain the narrow peak on Nov 20 as seen in the observations.}

>>>>

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. In the future, storm surge forecasts will make use of Neah Bay water level predictions from the National Oceanic and Atmospheric Administration (NOAA).  The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. Further, a geostrophic slope calculation with velocity 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca.}

@@

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error. Note that the model will contain tidal energy from shallow water constituents due to nonlinear interactions in the model equations and so the correction was calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (Sa, Ssa, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model sea surface height as a correction. Note that the harmonic analysis used to generate the constituents is applied in the same manner as that used in the forcing, i.e.\, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy before the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported in terms of five statistical measures of the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.  {\color{red} Observed and modelled maximum water levels and surge values as well as differences in their timing are reported in Table \ref{tab:peak}.}

@@

In general, there is also very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). {\color{red} The most noticeable difference is a broader peak in the model's surge curve, leading to a delay of a few hours in timing of the peak surge for most of the stations (Table \ref{tab:peak}). Note that the surge manifests as a broad peak in both the modelled and observed residual leading to a high probability that large residuals coincide with the high tide.} As such, it is of minor consequence that the timing of the observed and modelled maximum residuals do not match perfectly. {\color{red}Note that the observed surge curve has a relatively fast rise at each of the stations, a feature that is not reproduced by the model where the surge curves are generally symmetric. This sharp rise in sea level is not present in the remote forcing.  }

@@

\tbl{{\color{red}Summary of maximum water level, maximum surge amplitude and model delay for all simulations. The model time delay is the number of hours between the maximum in the observations and the maximum in the model. A negative number means the model's maximum was later than the observations. Values calculated over the same periods discussed in Table \ref{tab:statistics}. } }

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. This drop is smaller than the IB sea level change calculated by equation (\ref{eq:inverse}), approximately 1 cm rise in sea level associated with 1 hPa drop in atmospheric pressure. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This magnitude is in line with the findings of \citet{murty1995storm} that the IB contributes up to one third of the total surge amplitude. In our model, including the influence of remote atmospheric pressure forcing (in the sea level at the open boundary) produces realistic surge amplitudes compared with the runs only including local atmospheric forcing. Our results suggest that the local atmospheric pressure generates a sea level change that is smaller than the IB response. {\color{red} It is the atmospheric pressure gradient that drives changes in the sea surface height. Since our domain is relatively small, so is the atmospheric pressure gradient, and hence, we see only a small contribution from local atmospheric pressure. }

@@

In a simulation without atmospheric forcing, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side (Figure \ref{fig:spatial_sshonly}). While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland {\color{red}(Figure \ref{fig:spatial})}.  {\color{red} Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps because the incoming surge wave could be partially reflected from the mainland coast in the Strait of Georgia. Since local winds are stronger in this case, their contribution to the surge amplitude of about 6 cm at Point Atkinson is more significant than in the Feb 2006 case. However, this is still less than 10 \% of the total surge amplitude. }

@@

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind events on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was {\color{red}71.8 cm} which is slightly higher than that observed at Point Atkinson {\color{red}(70 cm)}. Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. The model results do not reflect a higher surge at Victoria {\color{red}(54 cm versus 58.7~cm at Point Atkinson)}, perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.  {\color{red} There is a fairly significant delay in the timing of the maximum surge at most stations (Table \ref{tab:peak}). The model's surge curve is much broader and does not contain the narrow peak on Nov 20 as seen in the observations. Both the observations and remote forcing contain a relatively fast rise in the surge curve that is not reproduced in the model, which might be explained by errors in the model wind speed and direction. }
